---
title: "DATA534_HW1"
author: "Oceanus Zhang"
date: "2024-02-01"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data 534 Homework 1

### Question 1
Consider the multiplication between a matrix

(a) What is the dimension of the resulting matrix C = A × B?

The dimension of the resulting matrix C=A×B will be a 4×4 matrix. This is because Matrix A has 4 rows and 3 columns, and Matrix B has 3 rows and 4 columns. The inner dimensions (3 columns of A and 3 rows of 

(b) Please show how you may compute the value of C2,2 using specific columns or rows from A and B.

To compute the value of C(2,2) (the element in the second row and second column of matrix C), I can take the dot product of the second row of Matrix A with the second column of Matrix B.
C(2,2)=A(2,1)\*B(1,2)+A(2,2)\*B(2,2)+A(2,3)\*B(3,2)=(5\*3)+(1\*1)+(2\*1)=18
 
(c) Show how you will define/declare the matrices A and B in R, and compute the matrix
C = A × B. 
```{r}
# Matrix A
matrix_a <- matrix(c(4, 2, 2,
                     5, 1, 2,
                     -1, 5, 1,
                     2, -3, 3), 
                   nrow = 4, byrow = TRUE)
# Matrix B
matrix_b <- matrix(c(2, 3, 1, 3,
                     4, 1, -3, 1,
                     1, 1, -1, 1), 
                   nrow = 3, byrow = TRUE)
# C=A*B
matrix_c <- matrix_a %*% matrix_b

# Print matrix C
print(matrix_c)

```

### Question 2
Please write a while loop to perform the following simple task: Start with the number 10 as the current value, at each step adds a random number between 0 and 1 to the current value; stop the loop until the new value is greater than or equal to 15
```{r }
# Initialize the current value
current_value <- 10

# Start the while loop
while (current_value < 15) {
  # Add a random number between 0 and 1 to the current value
  current_value <- current_value + runif(1, 0, 1)
  
  # Optionally, print the current value at each step
  print(current_value)
}

# Print the final result
print(paste("The final value is", current_value))
```

### Question 3
Consider a linear model defined by the function below, where a is a model parameter and x is input data. We assume that a = −0.2 and x follows a normal distribution with mean 0 and standard deviation 1.

output = function(a,x){
y = -1 + a*x
return(y)
}

(a) Generate a 5000 random samples of x and plot them on a histogram. 
```{r}
# Define the model parameter 'a'
a <- -0.2

# Define the linear model function
output <- function(a, x) {
  y <- -1 + a * x
  return(y)
}

# a) Generate 5000 random samples of x from a normal distribution with mean 0 and sd 1 and plot a histogram
x <- rnorm(5000, mean = 0, sd = 1)
hist(x, breaks = 30, col = 'navy', main = 'Histogram of x samples', xlab = 'x value', ylab = 'Frequency')
```

(b) Use the defined function to generate a set of y values given the model parameter a and the generated input data x. Plot y against x. 
```{r}
# (b) Generate a set of y values using the defined function and the generated x data and plot y agaist x
y <- output(a, x)
plot(x, y, main = 'Plot of y against x', xlab = 'x value', ylab = 'y value', pch = 19, col = 'orange', cex = 0.5)

```

### Question 4
This question follows Q3. In practice we do not know the value of a. To find the value of a, we first define the so-called log-likelihood function for this particular model (see below, you do not need to understand why it is defined in this way for now). The value of the (log)-likelihood measures the “fit” between the data (x and y) and the model given a particular value of model parameter (note that the input/argument x and y can be vectors, i.e., logL is a vectorized function)

(log-likelihood function)
logL = function(a, x, y){
sigma=1
500\*log(1/(sqrt(2\*pi)\*sigma)) - 1/(sigmaˆ2)\*(1/2)\*sum((a*x - y)ˆ2)
}

(a) Compute and plot the values of the log-likelihood function at different values of the modelparameter a specified in a.values defined below. Also, by visually inspecting the plot, roughly, what is the value of a on the x-axis that corresponds to the largest value of the log-likelihood? Use the same x and y generated from Q3.

(the range of the model parameter value to be considered)
a.values = seq(-2,2,0.1)

```{r }
### a)
# Define the log-likelihood function
logL <- function(a, x, y) {
  sigma <- 1
  n <- length(y)
  summand <- sum((y + 1 - a * x)^2)
  return(n * log(1 / (sqrt(2 * pi) * sigma)) - 1 / (2 * sigma^2) * summand)
}

# Compute the values of the log-likelihood function
a.values <- seq(-2, 2, 0.1)
logL_values <- sapply(a.values, function(a) logL(a, x, y))

# Plot the log-likelihood values
plot(a.values, logL_values, type = 'l', main = "Log-Likelihood Function", xlab = "a values", ylab = "Log-Likelihood")

```

(b) The answer in 4(a) is called the maximum likelihood estimator (MLE). Instead of doing this manually by trying all possible values of a, R has handy tools that can help us estimate the MLE numerically and
efficiently. You may use the R built-in function optim to find the MLE (reading the R help manual by typing “?optim” in R to understand this function yourself):

optim(par=0, logL, x=x, y=y,method = "L-BFGS-B",control = list(fnscale=-1))

The function argument f nscale is set to be -1, will this setting lead to minimization or maximization of the defined log-likelihood? 
```{r}
# Use the optim function to find the MLE
mle_results <- optim(par = 0, fn = logL, x = x, y = y, method = "L-BFGS-B", control = list(fnscale = -1))

```
(c) Set the initial value for the parameter to be 0.5, show the syntax to extract the MLE and report the MLE value. Is the value consistent to the answer in 4(a)? 
```{r}
# Set initial value and find the MLE
mle_results <- optim(par = 0.5, fn = logL, x = x, y = y, method = "L-BFGS-B", control = list(fnscale = -1))

# Extract the MLE value
mle_value <- mle_results$par

# Report the MLE value
mle_value
```


